\begin{acknowledgement}

    First I would like to thank my advisor, Professor Lars Ailo Bongo for his
    relentless support and encouragement during my time as a PhD student. He has
    indeed shown me what tough love is, and I think that I am grateful for that. 

    I would like to thank my co-advisors Professor Eiliv Lund and Associate
    Professor Karia Standahl Olsen for their wonderful ideas and warm welcome
    into a research field that was not my own. 

    I would like to extend my gratitude to Professor Michael Hallett and Vanessa
    Dumeaux for their hospitality when I visited their lab in Montreal in
    2016. I do not think this thesis would have been as interesting, without the
    projects I was fortunate enough to be a part of. Thank you! 

    I would like to thank my long time office wife Einar Holsbø, Morten
    Grønnesby, Nina Angelvik, and the BDPS lab at UiT. 

    Thank you to past or current students at UiT: Jan-Ove 'Kuken' Karlberg,
    Vegard Sandengen, Helge Hoff, Magnus Stenhaug, Erlend Graff, Kristian
    Elsebø, Martin Ernstsen, Amund Jensen, Michael Kampffmeyer, to name a few.
    You have all contributed to why I have been at the University for almost a
    decade.
    
    I would like to thank my colleagues at the Department of Computer Science,
    especially the technical staff, led by Maria Wulff Hauglann, down the hall
    from my old office. Also, the administrative staff at our School lab for
    their wonderful collaboration. 

    Thank you to the PhD students at Nordlandssykehuset in Bodø who have been my
    closest colleagues during the final push of my PhD. 
    
    I would like to thank my mom and dad, and my younger brother for their
    ever-present support. 

    Finally, Ane Sætrum for her continuous love and support, and her
    \emph{endurance} through all of my big or small projects. 

    % \begin{flushright}
    %     Bjørn Fjukstad, \\ 
    %     Bodø, June 2018. 
    % \end{flushright} 



\end{acknowledgement}


